On the specific recursive structure of iterated parametricity

### 2025-09-21 paper/restruct-new.tex
Update

@@

The well-foundedness of the construction can be justified by the observation that:
\begin{enumerate}
  \item Defining $\myframe^{n,p}$ for $p \leq n$ requires the definition
        of $\painting^{n-1,p-1}$, thus also $\myframe^{n-1,p-1}$ by typing
        dependency, as well as $\restrframe^{n,p-1}$, where
        $\restrframe^{n,p-1}$ in turn depends in $\myframe^{n,p-1}$ for its
        typing.

@@

        $$
          \begin{array}{llcl}
            \myframetype^{n,[0,...,p]}  &                                                     & :      & Type_{l+1}                                                               \\
            \myframetype^{n,[]}         &                                                     & \defeq & unit                                                                     \\
            \myframetype^{n,[0,...,p]}  &                                                     & \defeq & \myframetype^{n,[0,...,p-1]} \times Type_l                               \\
            \\
            \paintingtype^{n,[0,...,p]} & (\myframe^{n,[0,...,p]}:\myframetype^{n,[0,...,p]}) & :      & Type_{l+1}                                                               \\
            \paintingtype^{n,[]}        & \star                                               & \defeq & unit                                                                     \\
            \paintingtype^{n,[0,...,p]} & (\myframe^{n,[0,...,p]})                            & \defeq & \paintingtype^{n,[0,...,p-1]} \times (\myframe^{n,p} \rightarrow Type_l) \\
          \end{array}
        $$

@@

                        \restrframe^{n,[0,...,p-1]}:\restrframetype^{n,[0,...,p-1]}                             \\

@@

@@

@@

  \item Defining $\cohpainting^{n+1,p}$ for $p\leq n-1$, in addition to
        the type dependency in $\restrpainting^{n+1,p}$,
        $\restrpainting^{n,p}$, $\painting^{n+1,p}$, $\painting^{n,p}$,
        $\painting^{n-1,p}$, $\restrframe^{n+1,p}$, $\restrframe^{n,p}$,
        $\myframe^{n+1,p}$, $\myframe^{n,p}$ and $\myframe^{n-1,p}$ requires the
        definition of $\cohpainting^{n,k}$, thus also $\cohframe^{n,k}$ by
        typing dependency, for also $p \geq k$. By reasoning in sets, no
        higher-dimensional coherence frame is needed:
        \begin{center}
          \begin{tikzcd}
            & & painting^{n+1,p} \arrow[d, dotted] \\
            & & frame^{n+1,p} \\
            & & painting^{n,p} \arrow[d, dotted] & \restrpainting^{n+1,p} \arrow[d, dotted] \arrow[uul, dotted] \arrow[ul, dotted] \arrow[l, dotted] \\
            & & frame^{n,p} & \restrframe^{n+1,p} \arrow[uul, dotted] \arrow[l, dotted] \\
            & & painting^{n-1,p} \arrow[d, dotted] & \restrpainting^{n,p} \arrow[d, dotted] \arrow[uul, dotted] \arrow[ul, dotted] \arrow[l, dotted] & \cohpainting^{n+1,p} \arrow[ddllll] \arrow[l, dotted] \arrow[uul, dotted] \arrow[d, dotted]\\
            & & frame^{n-1,p} & \restrframe^{n,p} \arrow[l, dotted] & \cohframe^{n+1,p} \arrow[l, dotted] \arrow[uul, dotted]\\
            cohpainting^{n,k\geq p} \arrow[d, dotted] \\
            cohframe^{n,k\geq p} \\
          \end{tikzcd}
        \end{center}
\end{enumerate}

Looking at the dependencies in the reverse order, it is thus enough:
\begin{enumerate}
  \item To assume $\myframe^{n-2,k}$ and $\painting^{n-2,k}$ for all $k< p$
        in order to mutually define, for all $p\leq n$, the type of
        $\restrframe^{n-1,p}$ and $\myframe^{n-1,p}$ as a function from all
        $\restrframe^{n-1,k}$ for $k< p$. Using red for assumptions, this
        can be pictured as:
        \begin{center}
          \begin{tikzcd}
            \myframe^{n-1,p}(\textcolor{red}{restrframe^{n-1,k<p}}) \arrow[d] \\
            \textcolor{red}{painting^{n-2,k<p}} \arrow[d, dotted] \\
            \textcolor{red}{frame^{n-2,k<p}} \\
          \end{tikzcd}
        \end{center}

@@

        \begin{tiny}
          \begin{tikzcd}
            & & & frame^{n+1,p} \arrow[d] \arrow[ddr] \\
            & & & painting^{n,p} \arrow[ddlll] \arrow[dddr] \arrow[d, dotted]\\
            & & & frame^{n,p} \arrow[d] \arrow[ddr] & \restrframe^{n+1,k<p} \arrow[d] \arrow[ddr] \arrow[l, dotted] \arrow[uul, dotted, shift right=2mm] \\
            \painting^{n-1,k\geq p} \arrow[d, dotted] \arrow[dddr] \arrow[dd, bend right=70] & & &
            painting^{n-1,p} \arrow[ddlll] \arrow[dddll] \arrow[d, dotted] & \restrpainting^{n,k<p} \arrow[d, dotted] \arrow[ddlll] \arrow[dddll] \arrow[l, dotted] &
            \cohpainting^{n+1,p} \arrow[ddlll] \arrow[d, dotted] \arrow[l, dotted] \\
            \myframe^{n-1,k\geq p} \arrow[d] \arrow[ddr] & \restrframe^{n,k\geq p} \arrow[d] \arrow[ddr] \arrow[l, dotted] & &
            frame^{n-1,p} \arrow[d] \arrow[ddr] & \restrframe^{n,k<p} \arrow[d] \arrow[ddr] \arrow[l, dotted] \arrow[uul, dotted, shift right=2mm] &
            \cohframe^{n+1,p} \arrow[d] \arrow[l, dotted] \arrow[uul, dotted, shift right=2mm]\\
            \textcolor{red}{painting^{n-2,k\geq p}} \arrow[d, dotted] &
            \textcolor{red}{restrpainting^{n-1,k\geq p}} \arrow[l, dotted] \arrow[d, dotted] &
            \textcolor{red}{cohpainting^{n,k\geq p}} \arrow[l, dotted] \arrow[d, dotted] &
            \textcolor{red}{painting^{n-2,k<p}} \arrow[d, dotted] &
            \textcolor{red}{restrpainting^{n-1,k<p}} \arrow[l, dotted] \arrow[d, dotted] &
            \textcolor{red}{cohpainting^{n,k<p}} \arrow[l, dotted] \arrow[d, dotted] \\
            \textcolor{red}{frame^{n-2,k\geq p}} &
            \textcolor{red}{restrframe^{n-1,k\geq p},X_{n-2}} \arrow[l, dotted] &
            \textcolor{red}{cohframe^{n,k\geq p},X_{n-1}} \arrow[l, dotted] &
            \textcolor{red}{frame^{n-2,k<p}} &
            \textcolor{red}{restrframe^{n-1,k<p}} \arrow[l, dotted] \arrow[uul, dotted, shift right=2mm] &
            \textcolor{red}{cohframe^{n,k<p}} \arrow[l, dotted] \arrow[uul, dotted, shift right=2mm] \\
          \end{tikzcd}
        \end{tiny}
\end{enumerate}

>>>>

\title{On the specific recursive structure of iterated parametricity}

\maketitle

\begin{abstract}
  In a previous paper, we gave equations describing the construction
  of semi-simplicial and semi-cubical sets via iterated parametricity.
  The construction was formalised in the Rocq (ex-Coq) proof assistant
  building $n$-truncated sets using an induction on a large structure
  embedding equational reasoning.

  The current paper describes a new and lighter presentation of the
  construction where equational reasoning is replaced by
  definition. Being lighter, it can now be described in full details,
  closely following the corresponding new machine-checked
  formalisation.
\end{abstract}

@@

The well-foundedness of the construction can be justified by the observation that:
\begin{itemize}
  \item Defining $\myframe^{n,p}$ for $p \leq n$ requires the definition
        of $\painting^{n-1,p-1}$, thus also $\myframe^{n-1,p-1}$ by typing
        dependency, as well as $\restrframe^{n,p-1}$, where
        $\restrframe^{n,p-1}$ in turn depends in $\myframe^{n,p-1}$ for its
        typing.

@@

        $$
          \begin{array}{llcl}
            \myframetype^{n,[0,...,p]}  &                                                     & :      & Type_{l+1}                                                               \\
            \myframetype^{n,[]}         &                                                     & \defeq & unit                                                                     \\
            \myframetype^{n,[0,...,p]}  &                                                     & \defeq & \myframetype^{n,[0,...,p-1]} \times Type_l                               \\
            \\
            \paintingtype^{n,[0,...,p]} & (\myframe^{n,[0,...,p]}:\myframetype^{n,[0,...,p]}) & :      & Type_{l+1}                                                               \\
            \paintingtype^{n,[]}        & \star                                               & \defeq & unit                                                                     \\
            \paintingtype^{n,[0,...,p]} & (\myframe^{n,[0,...,p]})                            & \defeq & \paintingtype^{n,[0,...,p-1]}(\myframe^{n,[0,...,p-1]}) \times (\myframe^{n,p} \rightarrow Type_l) \\
          \end{array}
        $$

@@

                        \restrframe^{n,[0,...,p-1]}:\restrframetype^{n,[0,...,p-1]}

@@

        Or, alternatively, as:
        $$\!\!\!\!\!\!\!\!\!\!\!\!
                      \end{array}\right)(\restrframe^{n,[0,...,p-1]}) \\
                      \end{array}\right) : \restrframetype^{n,[0,...,p-1]}
                        \end{array}\right) \rightarrow \myframetype^{n-1,[0,...,p]}
                    \end{array}\right)                                     & \defeq & \star \mapsto (\star,unit)      \\
            \restrframe^{n,[0,...,p]} \mapsto
                        \end{array}\right)(\restrframe^{n,[0,...,p-1]}), \\
                        \end{array}\right)(\restrframe^{n,[0,...,p-1]}).\, \\

        Or, even, using an alternative notation, as:
        $$\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!
          \begin{array}{llcl}
            \restrframetype^{n,<p}                            &
            \multicolumn{3}{l}{
              \left(\begin{array}{l}
                        \myframe^{n-1,<p}:\myframetype^{n-1,<p}                               \\
                        \painting^{n-1,<p}:\paintingtype^{n-1,<p}(\myframe^{n-1,<p}) \\
                      \end{array}\right) : Type_{l}
            }                                                                                         \\
            \restrframetype^{n,<0}                                     &
            \left(\begin{array}{l}
                      \star \\
                      \star \\
                    \end{array}\right)                                     & \defeq &
            unit                                                                                      \\
            \restrframetype^{n,\leq p}                              &
            \left(\begin{array}{l}
                      \myframe^{n-1,\leq p}  \\
                      \painting^{n-1,\leq p} \\
                    \end{array}\right) & \defeq &
            \begin{array}{l}
              \Sigma \restrframe^{n,<p}                            \\
              \quad:\restrframetype^{n,<p}
              \left(\begin{array}{l}
                        \myframe^{n-1,<p}  \\
                        \painting^{n-1,<p} \\
                      \end{array}\right)
              .\,                                                           \\
              \Pi q\leq n.\,\Pi \omega.\,\myframe^{n,\leq p}
              \left(\begin{array}{l}
                        \myframe^{n-1,<p}  \\
                        \painting^{n-1,<p} \\
                      \end{array}\right)(\restrframe^{n,<p}) \\
            \end{array} \\
            \myframe^{n,\leq p}                                     &
            \multicolumn{3}{l}{
              \left(\begin{array}{l}
                        \myframe^{n-1,<p}:\myframetype^{n-1,<p}                               \\
                        \painting^{n-1,<p}:\paintingtype^{n-1,<p}(\myframe^{n-1,<p}) \\
                      \end{array}\right) : \restrframetype^{n,<p}
                        \left(\begin{array}{l}
                          \myframe^{n-1,<p}  \\
                          \painting^{n-1,<p} \\
                        \end{array}\right) \rightarrow \myframetype^{n-1,\leq p}
            }                                                                                         \\
            \myframe^{n,\leq 0}                                           &
            \left(\begin{array}{l}
                      \star \\
                      \star \\
                    \end{array}\right)                                     & \defeq & \star \mapsto (\star,unit)      \\
            \myframe^{n,\leq p+1}                                   &
            \left(\begin{array}{l}
                      \myframe^{n-1,\leq p}  \\
                      \painting^{n-1,\leq p} \\
                    \end{array}\right) & \defeq &
            \restrframe^{n,\leq p} \mapsto
            \left(\begin{array}{l}
                      \myframe^{n,\leq p}
                      \left(\begin{array}{l}
                          \myframe^{n-1,<p}  \\
                          \painting^{n-1,<p} \\
                        \end{array}\right)(\restrframe^{n,<p}), \\
                      \Sigma d:\myframe^{n,\leq p}
                      \left(\begin{array}{l}
                          \myframe^{n-1,<p}  \\
                          \painting^{n-1,<p} \\
                        \end{array}\right)(\restrframe^{n,<p}).\, \\
                      \quad \Pi\omega.\,\painting^{n-1,p}(\restrframe^{n,p}(d))     \\
                    \end{array}\right)                    \\
          \end{array}
        $$

@@

To define $\painting$ formally, we need to define $\myframe^{n,[p,...,n]}$, $\painting^{n,[p,...,n]}$ and $\restrframe^{n,[p,...,n]}$:
\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!\!
            \myframetype^{n,[p,...,n]}  &                                                     & :      & Type_{l+1}                                                               \\
            \myframetype^{n,[]}         &                                                     & \defeq & unit                                                                     \\
            \myframetype^{n,[p,...,n]}  &                                                     & \defeq & Type_l \times \myframetype^{n,[p+1,...,n-1]}                               \\
            \\
            \paintingtype^{n,[p,...,n]} & (\myframe^{n,[p,...,n]}:\myframetype^{n,[p,...,n]}) & :      & Type_{l+1}                                                               \\
            \paintingtype^{n,[]}        & \star                                               & \defeq & unit                                                                     \\
            \paintingtype^{n,[p,...,n]}_{p<n} & (\myframe^{n,[p,...,n]})                            & \defeq & (\myframe^{n,p} \rightarrow Type_l) \times \paintingtype^{n,[p+1,...,n]}(\myframe^{n,[p+1,...,n]}) \\
            \\
            \restrframetype^{n,[p,...,n]} &
                        \myframe^{n-1,[p,...,n-1]}:\myframetype^{n-1,[p,...,n-1]}                                     \\
                        \painting^{n-1,[p,...,n-1]}:\paintingtype^{n-1,[p,...,n-1]}(\myframe^{n-1,[p,...,n-1]})           \\
                        \restrframe^{n,[0,...,p-1]}:\restrframetype^{n,[0,...,p-1]}
                      \myframe^{n-1,[0,...,n-1]} \\
                      \painting^{n-1,[0,...,n-1]} \\
                      \restrpainting^{n,[0,...,n-1]} \\
                    \end{array}\right)                                     & \defeq & unit \\
            \restrframetype^{n,[p,...,n-1]} &
                                    \myframe^{n-1,[p,...,n-1]}  \\
                                    \painting^{n-1,[p,...,n-1]} \\
            & \defeq & 
                                  \myframe^{n,p}
                                \rightarrow \myframe^{n-1,p}\\
                                \times \\
                                \restrframetype^{n,[p+1,...,n]}
                                    \myframe^{n-1,[p+1,...,n-1]}  \\
                                    \painting^{n-1,[p+1,...,n-1]} \\

        This allows to define $\painting^{n,p}$ formally:
        $$
          \begin{array}{llcl}
            \painting^{n,p}                                     &
            \multicolumn{3}{l}{
              \left(\begin{array}{l}
                        \myframe^{n-1,[0,...,p-1]}:\myframetype^{n-1,[0,...,p-1]}                               \\
                        \painting^{n-1,[0,...,p-1]}:\paintingtype^{n-1,[0,...,p-1]}(\myframe^{n-1,[0,...,p-1]}) \\
                        \myframe^{n-1,[p,...,n-1]}:\myframetype^{n-1,[p,...,n-1]}                                     \\
                        \painting^{n-1,[p,...,n-1]}:\paintingtype^{n-1,[p,...,n-1]}(\myframe^{n-1,[p,...,n-1]})           \\
                        \restrframe^{n,[0,...,p-1]}:\restrframetype^{n,[0,...,p-1]}
                                \left(\begin{array}{l}
                                    \myframe^{n-1,[0,...,p-1]}  \\
                                    \painting^{n-1,[0,...,p-1]} \\
                                  \end{array}\right) \\
                        \restrframe^{n,[p,...,n-1]}:\restrframetype^{n,[p,...,n-1]}
                                \left(\begin{array}{l}
                                    \myframe^{n-1,[0,...,p-1]}  \\
                                    \painting^{n-1,[0,...,p-1]} \\
                                    \myframe^{n-1,[p,...,n-2]}  \\
                                    \painting^{n-1,[p,...,n-2]} \\
                                    \restrframe^{n,[0,...,p-1]} \\
                                  \end{array}\right) \\
                        E:\myframe^{n-1,n-1}\rightarrow Type_l\\
                      \end{array}\right) : \myframe^{n-1,n-1}\rightarrow Type_l
            }    \\
            \painting^{n,n}                                           &
            \left(\begin{array}{l}
                      \myframe^{n-1,[0,...,n-1]} \\
                      \painting^{n-1,[0,...,n-1]} \\
                      \star \\
                      \star \\
                      \restrpainting^{n,[0,...,n-1]} \\
                      \star \\
                      E \\
                    \end{array}\right)                                     & \defeq & E \\
            \painting^{n,p}_{p<n}                                   &
                                \left(\begin{array}{l}
                                    \myframe^{n-1,[0,...,p-1]}  \\
                                    \painting^{n-1,[0,...,p-1]} \\
                                    \myframe^{n-1,[p,...,n-1]}  \\
                                    \painting^{n-1,[p,...,n-1]} \\
                                    \restrframe^{n,[0,...,p-1]} \\
                                    \restrframe^{n,[p,...,n-1]} \\
                                    E \\
                                  \end{array}\right) & \defeq & d \mapsto
            \left(\begin{array}{l}
              \Sigma l:(\Pi q\leq n.\,\Pi \omega.\
                      \painting^{n-1,[0,...,p]}(\restrframe^{n,p}(d))). \\
                      \painting^{n,[p+1,...,n-1]}
                      \left(\begin{array}{l}
                                    \myframe^{n-1,[0,...,p]}  \\
                                    \painting^{n-1,[0,...,p]} \\
                                    \myframe^{n-1,[p+1,...,n-1]}  \\
                                    \painting^{n-1,[p+1,...,n-1]} \\
                                    \restrframe^{n,[0,...,p]} \\
                                    \restrframe^{n,[p+1,...,n-1]} \\
                                    E \\
                        \end{array}\right)(d,l) \\
                    \end{array}\right)                    \\
          \end{array}
          $$

          From which we can deduce $\painting^{n,[0,...,p]}$ by recursion on $p$.

@@

  \item Defining $\cohpainting^{n+1,p}$ for $p\leq n-1$, in addition to
        the type dependency in $\restrpainting^{n+1,p}$,
        $\restrpainting^{n,p}$, $\painting^{n+1,p}$, $\painting^{n,p}$,
        $\painting^{n-1,p}$, $\restrframe^{n+1,p}$, $\restrframe^{n,p}$,
        $\myframe^{n+1,p}$, $\myframe^{n,p}$ and $\myframe^{n-1,p}$ requires the
        definition of $\cohpainting^{n,k}$, thus also $\cohframe^{n,k}$ by
        typing dependency, for also $p \geq k$. By reasoning in sets, no
        higher-dimensional coherence frame is needed:
        \begin{center}
          \begin{tikzcd}
            & & painting^{n+1,p} \arrow[d, dotted] \\
            & & frame^{n+1,p} \\
            & & painting^{n,p} \arrow[d, dotted] & \restrpainting^{n+1,p} \arrow[d, dotted] \arrow[uul, dotted] \arrow[ul, dotted] \arrow[l, dotted] \\
            & & frame^{n,p} & \restrframe^{n+1,p} \arrow[uul, dotted] \arrow[l, dotted] \\
            & & painting^{n-1,p} \arrow[d, dotted] & \restrpainting^{n,p} \arrow[d, dotted] \arrow[uul, dotted] \arrow[ul, dotted] \arrow[l, dotted] & \cohpainting^{n+1,p} \arrow[ddllll] \arrow[l, dotted] \arrow[uul, dotted] \arrow[d, dotted]\\
            & & frame^{n-1,p} & \restrframe^{n,p} \arrow[l, dotted] & \cohframe^{n+1,p} \arrow[l, dotted] \arrow[uul, dotted]\\
            cohpainting^{n,k\geq p} \arrow[d, dotted] \\
            cohframe^{n,k\geq p} \\
          \end{tikzcd}
        \end{center}
\end{itemize}

Looking at the dependencies in the reverse order, it is thus enough:
\begin{itemize}
  \item To assume $\myframe^{n-2,k}$ and $\painting^{n-2,k}$ for all $k< p$
        in order to mutually define, for all $p\leq n$, the type of
        $\restrframe^{n-1,p}$ and $\myframe^{n-1,p}$ as a function from all
        $\restrframe^{n-1,k}$ for $k< p$. Using red for assumptions, this
        can be pictured as:
        \begin{center}
          \begin{tikzcd}
            \myframe^{n-1,p}(\textcolor{red}{restrframe^{n-1,k<p}}) \arrow[d] \\
            \textcolor{red}{painting^{n-2,k<p}} \arrow[d, dotted] \\
            \textcolor{red}{frame^{n-2,k<p}} \\
          \end{tikzcd}
        \end{center}

@@

        \begin{tiny}
          \begin{tikzcd}
            & & & frame^{n+1,p} \arrow[d] \arrow[ddr] \\
            & & & painting^{n,p} \arrow[ddlll] \arrow[dddr] \arrow[d, dotted]\\
            & & & frame^{n,p} \arrow[d] \arrow[ddr] & \restrframe^{n+1,k<p} \arrow[d] \arrow[ddr] \arrow[l, dotted] \arrow[uul, dotted, shift right=2mm] \\
            \painting^{n-1,k\geq p} \arrow[d, dotted] \arrow[dddr] \arrow[dd, bend right=70] & & &
            painting^{n-1,p} \arrow[ddlll] \arrow[dddll] \arrow[d, dotted] & \restrpainting^{n,k<p} \arrow[d, dotted] \arrow[ddlll] \arrow[dddll] \arrow[l, dotted] &
            \cohpainting^{n+1,p} \arrow[ddlll] \arrow[d, dotted] \arrow[l, dotted] \\
            \myframe^{n-1,k\geq p} \arrow[d] \arrow[ddr] & \restrframe^{n,k\geq p} \arrow[d] \arrow[ddr] \arrow[l, dotted] & &
            frame^{n-1,p} \arrow[d] \arrow[ddr] & \restrframe^{n,k<p} \arrow[d] \arrow[ddr] \arrow[l, dotted] \arrow[uul, dotted, shift right=2mm] &
            \cohframe^{n+1,p} \arrow[d] \arrow[l, dotted] \arrow[uul, dotted, shift right=2mm]\\
            \textcolor{red}{painting^{n-2,k\geq p}} \arrow[d, dotted] &
            \textcolor{red}{restrpainting^{n-1,k\geq p}} \arrow[l, dotted] \arrow[d, dotted] &
            \textcolor{red}{cohpainting^{n,k\geq p}} \arrow[l, dotted] \arrow[d, dotted] &
            \textcolor{red}{painting^{n-2,k<p}} \arrow[d, dotted] &
            \textcolor{red}{restrpainting^{n-1,k<p}} \arrow[l, dotted] \arrow[d, dotted] &
            \textcolor{red}{cohpainting^{n,k<p}} \arrow[l, dotted] \arrow[d, dotted] \\
            \textcolor{red}{frame^{n-2,k\geq p}} &
            \textcolor{red}{restrframe^{n-1,k\geq p},X_{n-2}} \arrow[l, dotted] &
            \textcolor{red}{cohframe^{n,k\geq p},X_{n-1}} \arrow[l, dotted] &
            \textcolor{red}{frame^{n-2,k<p}} &
            \textcolor{red}{restrframe^{n-1,k<p}} \arrow[l, dotted] \arrow[uul, dotted, shift right=2mm] &
            \textcolor{red}{cohframe^{n,k<p}} \arrow[l, dotted] \arrow[uul, dotted, shift right=2mm] \\
          \end{tikzcd}
        \end{tiny}
\end{itemize}
